
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Loading data and running QC

Load packages, set working directory and random seed.

```{r results='hide', message=FALSE, warning=FALSE}
require(dada2)
require(phyloseq)
require(Biostrings)
require(ggplot2)
require(DECIPHER)   
require(phangorn)   # for phylogenetic tree construction
require(ape)        # tree manipulation / I/O

set.seed(42)
setwd('./')
```

```{r}
path = "./"

forward.raw <- sort(list.files(path, pattern = "_R1.*\\.fq", full.names = TRUE))
reverse.raw <- sort(list.files(path, pattern = "_R2.*\\.fq", full.names = TRUE))

# must be same number 
stopifnot(length(forward.raw) == length(reverse.raw))

# extract names
sample.names <- sapply(basename(forward.raw),
                       function(x) sub("_R1.*", "", x))
sample.names <- unname(sample.names)
sample.names
```

```{r warning=FALSE}
plotQualityProfile(forward.raw, n = 50000)
plotQualityProfile(reverse.raw, n = 50000)
```

```{r warning=FALSE}
plotQualityProfile(forward.raw, aggregate = TRUE, n = 100000)
plotQualityProfile(reverse.raw, aggregate = TRUE, n = 100000)
```

## Cutadapt

```{r}
path <- "./trimmed_reads"
forward.cut <- sort(list.files(path, pattern = "_R1\\.fq\\.gz$", full.names = TRUE))
reverse.cut <- sort(list.files(path, pattern = "_R2\\.fq\\.gz$", full.names = TRUE))
```

```{r}
cutadapt_summary <- read.table("trimmed_reads/cutadapt_summary.tsv",
                               header = TRUE, sep = "\t")
```
## Error model

Train a parametric error model **separately** for forward and reverse reads
(they have different error profiles).

```{r}
forward.err <- learnErrors(forward.qc, multithread = TRUE, MAX_CONSIST = 20)
reverse.err <- learnErrors(reverse.qc, multithread = TRUE, MAX_CONSIST = 20)
```

```{r warning=FALSE}
plotErrors(forward.err, nominalQ = TRUE)
plotErrors(reverse.err, nominalQ = TRUE)
```

## Dereplication and DADA2 inference
```{r echo=T, results='hide', message=F, warning=FALSE}
forward.derep <- derepFastq(forward.qc, n = 1e7, verbose = TRUE)
reverse.derep <- derepFastq(reverse.qc, n = 1e7, verbose = TRUE)

names(forward.derep) <- sample.names
names(reverse.derep) <- sample.names

forward.dada <- dada(forward.derep, err = forward.err,
                     pool = FALSE, multithread = TRUE)
reverse.dada <- dada(reverse.derep, err = reverse.err,
                     pool = FALSE, multithread = TRUE)
```

## Merge paired reads
```{r echo=T, message=F, warning=FALSE}
mergers <- mergePairs(forward.dada, forward.derep,
                      reverse.dada, reverse.derep,
                      minOverlap = 12,
                      maxMismatch = 0,
                      verbose = TRUE)

head(mergers[[1]])

sequence.table <- makeSequenceTable(mergers)
dim(sequence.table)

table(nchar(getSequences(sequence.table)))
```

Optionally, restrict to the expected amplicon length range to remove obvious
non-target sequences:

```{r}
seqtab.lenfilt <- sequence.table[, nchar(colnames(sequence.table)) %in% 400:430]
dim(seqtab.lenfilt)
```

## Chimera removal

```{r}
seqtab.nochim <- removeBimeraDenovo(seqtab.lenfilt,
                                    method = "consensus",
                                    multithread = TRUE,
                                    verbose = TRUE)
dim(seqtab.nochim)
sum(seqtab.nochim) / sum(seqtab.lenfilt)
```

## Track reads through the pipeline

```{r}
getN <- function(x) sum(getUniques(x))
track <- cbind(qc.out,
               sapply(forward.dada, getN),
               sapply(reverse.dada, getN),
               sapply(mergers, getN),
               rowSums(seqtab.nochim))
colnames(track) <- c("input", "filtered",
                     "denoisedF", "denoisedR",
                     "merged", "nonchim")
rownames(track) <- sample.names
track

dir.create("results", showWarnings = FALSE)
write.csv(track, "results/track.csv", row.names = TRUE)
```

## Taxonomy assignment

```{r}
taxa <- assignTaxonomy(seqtab.nochim,
                       "./silva_nr_v132_train_set.fa.gz",
                       multithread = TRUE, tryRC = TRUE)
taxa <- addSpecies(taxa,
                   "./silva_species_assignment_v132.fa.gz",
                   tryRC = TRUE)
```

## Phylogenetic tree (DECIPHER + phangorn)



```{r message=FALSE, warning=FALSE}
# Pull ASV sequences from the chimera-free table
seqs <- getSequences(seqtab.nochim)
names(seqs) <- seqs   # required by phangorn downstream

# Multiple sequence alignment
alignment <- AlignSeqs(DNAStringSet(seqs),
                       anchor = NA,
                       processors = NULL,
                       verbose = FALSE)

# Convert to phangorn format
phang.align <- phyDat(as(alignment, "matrix"), type = "DNA")

# Distance matrix and NJ starting tree
dm     <- dist.ml(phang.align)
treeNJ <- NJ(dm)

# ML optimization under GTR+G+I
fit    <- pml(treeNJ, data = phang.align)
fitGTR <- update(fit, k = 4, inv = 0.2)
fitGTR <- optim.pml(fitGTR,
                    model         = "GTR",
                    optInv        = TRUE,
                    optGamma      = TRUE,
                    rearrangement = "stochastic",
                    control       = pml.control(trace = 0))

# Save the tree for reuse
dir.create("results", showWarnings = FALSE)
ape::write.tree(fitGTR$tree, file = "results/asv_tree.nwk")
```

## Build the phyloseq object (now with a tree)

Prepare metadata as before.

```{r}
metadata <- read.table(file = "sample-metadata.tsv",
                       comment.char = "", header = TRUE, sep = "\t")
manifest <- read.table(file = "manifest.tsv",
                       header = TRUE, sep = "\t")

manifest$absolute.filepath <- substring(manifest$absolute.filepath, 6)

merged_table   <- merge(manifest, metadata,
                        by.x = "sample.id", by.y = "X.SampleID")
resulting_table <- merged_table[, c(2, 1, 3:7)]
rownames(resulting_table) <- merged_table$absolute.filepath
resulting_table <- resulting_table[, -1]
```

Assemble the phyloseq object — this time including the tree.

```{r}
ps <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows = FALSE),
               sample_data(resulting_table),
               tax_table(taxa),
               phy_tree(fitGTR$tree))
```

Replace long DNA strings with short ASV IDs (DNA preserved in `refseq`).

```{r}
dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps <- merge_phyloseq(ps, dna)
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))
ps
```

Now phylogenetic metrics are available, e.g.:

```{r eval=FALSE}
# Faith's PD via picante (if installed)
# library(picante)
# pd <- picante::pd(as(otu_table(ps), "matrix"), phy_tree(ps))

# UniFrac distances
unifrac.uw <- UniFrac(ps, weighted = FALSE)
unifrac.w  <- UniFrac(ps, weighted = TRUE)
```

