---
title: "16S rRNA DADA2 pipeline (paired-end, interactive)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

Interactive, chunk-by-chunk equivalent of the Snakemake pipeline (see the
repo's `README.md` and `Snakefile`) for a single paired-end 16S run: primer
trimming, quality filtering, DADA2 denoising, chimera removal, taxonomy
(SILVA + RDP), an ASV abundance table, a composition plot, and a phylogenetic
tree, assembled into a `phyloseq` object. Run it from RStudio ("Run Cell") or
render it end to end with:

```
Rscript -e "rmarkdown::render('scripts/dada2_paired_with_tree.Rmd')"
```

Expects raw reads in the working directory matching `{sample}_R1*.fq*` /
`{sample}_R2*.fq*`, plus `sample-metadata.tsv` and `manifest.tsv` for the
`phyloseq` sample data. Reference databases and primer FASTAs are read from
the repository (`references/`, `primers/`) via `repo_root` below.

## Setup

Load packages, set the working directory, seed, and paths to the shared
repo resources (references / primers) so the rest of the notebook can use
them regardless of which run directory it's launched from.

```{r results='hide', message=FALSE, warning=FALSE}
require(dada2)
require(phyloseq)
require(Biostrings)
require(ggplot2)
require(tidyr)
require(DECIPHER)   # alignment + tree building
require(ggtree)      # tree visualization
require(ape)          # tree manipulation / I/O

set.seed(42)
setwd('./')

# Path to the pipeline repository root (contains scripts/, primers/,
# references/). Adjust if this notebook's working directory sits somewhere
# other than one level below the repo root.
repo_root  <- ".."
ref_dir    <- file.path(repo_root, "references")
primer_dir <- file.path(repo_root, "primers")
```

## Loading data and running QC

```{r}
path = "./"

forward.raw <- sort(list.files(path, pattern = "_R1.*\\.fq", full.names = TRUE))
reverse.raw <- sort(list.files(path, pattern = "_R2.*\\.fq", full.names = TRUE))

# must be same number
stopifnot(length(forward.raw) == length(reverse.raw))

# extract names
sample.names <- sapply(basename(forward.raw),
                       function(x) sub("_R1.*", "", x))
sample.names <- unname(sample.names)
sample.names
```

Per-sample quality profiles (subsampled to 50k reads each):

```{r warning=FALSE}
plotQualityProfile(forward.raw, n = 50000)
plotQualityProfile(reverse.raw, n = 50000)
```

Aggregated quality profiles across all samples (100k reads):

```{r warning=FALSE}
plotQualityProfile(forward.raw, aggregate = TRUE, n = 100000)
plotQualityProfile(reverse.raw, aggregate = TRUE, n = 100000)
```

## Primer trimming (cutadapt)

Strips primers with `scripts/run_cutadapt.sh` (the interactive counterpart
of the Snakemake `trim_primers` rule) before DADA2 ever sees the reads.
Run this once per dataset — re-running is safe but redundant, so it's
`eval=FALSE` by default; flip to `eval=TRUE` or run the chunk manually.

```{r cutadapt-run, eval=FALSE}
dir.create("trimmed_reads", showWarnings = FALSE)
system2("bash",
        c(file.path(repo_root, "scripts", "run_cutadapt.sh"),
          ".", "trimmed_reads",
          file.path(primer_dir, "fwd_primers.fasta"),
          file.path(primer_dir, "rev_primers.fasta")))
```

Load the trimmed reads and the per-sample summary it produced:

```{r}
path <- "./trimmed_reads"
forward.cut <- sort(list.files(path, pattern = "_R1\\.fq\\.gz$", full.names = TRUE))
reverse.cut <- sort(list.files(path, pattern = "_R2\\.fq\\.gz$", full.names = TRUE))
```

```{r}
cutadapt_summary <- read.table("trimmed_reads/cutadapt_summary.tsv",
                               header = TRUE, sep = "\t")
cutadapt_summary
```

## Filter and trim

`filterAndTrim` on the primer-trimmed reads — same parameters as
`scripts/filter_reads.R` (`truncLen = c(120, 120)`). Inspect the quality
profiles above and lower `truncLen` if reads are lost heavily here; R1 +
R2 truncated lengths must sum to more than amplicon length + required
overlap (see README Troubleshooting).

```{r}
filt_path <- "filtered"
forward.qc <- file.path(filt_path, paste0(sample.names, "_R1.fq.gz"))
reverse.qc <- file.path(filt_path, paste0(sample.names, "_R2.fq.gz"))
names(forward.qc) <- sample.names
names(reverse.qc) <- sample.names

qc.out <- filterAndTrim(forward.cut, forward.qc,
                        reverse.cut, reverse.qc,
                        truncLen    = c(120, 120),
                        maxN        = 0,
                        maxEE       = c(2, 2),
                        truncQ      = 2,
                        rm.phix     = TRUE,
                        compress    = TRUE,
                        multithread = TRUE)
qc.out
```

## Error model

Train a parametric error model **separately** for forward and reverse reads
(they have different error profiles).

```{r}
forward.err <- learnErrors(forward.qc, multithread = TRUE, MAX_CONSIST = 20)
reverse.err <- learnErrors(reverse.qc, multithread = TRUE, MAX_CONSIST = 20)
```

```{r warning=FALSE}
plotErrors(forward.err, nominalQ = TRUE)
plotErrors(reverse.err, nominalQ = TRUE)
```

## Dereplication and DADA2 inference
```{r echo=T, results='hide', message=F, warning=FALSE}
forward.derep <- derepFastq(forward.qc, n = 1e7, verbose = TRUE)
reverse.derep <- derepFastq(reverse.qc, n = 1e7, verbose = TRUE)

names(forward.derep) <- sample.names
names(reverse.derep) <- sample.names

forward.dada <- dada(forward.derep, err = forward.err,
                     pool = FALSE, multithread = TRUE)
reverse.dada <- dada(reverse.derep, err = reverse.err,
                     pool = FALSE, multithread = TRUE)
```

## Merge paired reads
```{r echo=T, message=F, warning=FALSE}
mergers <- mergePairs(forward.dada, forward.derep,
                      reverse.dada, reverse.derep,
                      minOverlap = 12,
                      maxMismatch = 0,
                      verbose = TRUE)

head(mergers[[1]])

sequence.table <- makeSequenceTable(mergers)
dim(sequence.table)

table(nchar(getSequences(sequence.table)))
```

### Length filtering (optional)

Restrict to the expected amplicon length range to remove obvious
non-target sequences. Set the range from the `table()` output above and the
expected length of the primer set actually used (V3-V4 ~460bp, V1-V2/V9
~250-300bp, etc.) — the 400:430 default below assumes a V3-V4-length
amplicon and should be adjusted otherwise. If you skip this chunk, use
`sequence.table` instead of `seqtab.lenfilt` in Chimera removal below.

```{r}
seqtab.lenfilt <- sequence.table[, nchar(colnames(sequence.table)) %in% 400:430]
dim(seqtab.lenfilt)
```

## Chimera removal

```{r}
seqtab.nochim <- removeBimeraDenovo(seqtab.lenfilt,
                                    method = "consensus",
                                    multithread = TRUE,
                                    verbose = TRUE)
dim(seqtab.nochim)
sum(seqtab.nochim) / sum(seqtab.lenfilt)
```

## Track reads through the pipeline

```{r}
getN <- function(x) sum(getUniques(x))
track <- cbind(qc.out,
               sapply(forward.dada, getN),
               sapply(reverse.dada, getN),
               sapply(mergers, getN),
               rowSums(seqtab.nochim))
colnames(track) <- c("input", "filtered",
                     "denoisedF", "denoisedR",
                     "merged", "nonchim")
rownames(track) <- sample.names
track

dir.create("results", showWarnings = FALSE)
write.csv(track, "results/track.csv", row.names = TRUE)
```

## Taxonomy assignment

### SILVA (genus-level + species)

Matches `scripts/assign_taxa.R` + `scripts/add_species.R`
(SILVA v138.2, from Zenodo — see README "Reference databases").

```{r}
taxa <- assignTaxonomy(seqtab.nochim,
                       file.path(ref_dir, "silva_nr99_v138.2_toSpecies_trainset.fa.gz"),
                       multithread = TRUE, tryRC = TRUE)
taxa <- addSpecies(taxa,
                   file.path(ref_dir, "silva_v138.2_assignSpecies.fa.gz"),
                   tryRC = TRUE)

dir.create("results/taxa/silva", showWarnings = FALSE, recursive = TRUE)
saveRDS(taxa, "results/taxa/silva/taxa_species.rds")
head(taxa)
```

### RDP (alternative, single-step)

Matches `scripts/assign_taxa.R` as used by the `assign_taxonomy_rdp` rule:
the RDP 19 training set already resolves to species, so no separate
`addSpecies` call is needed.

```{r}
taxa.rdp <- assignTaxonomy(seqtab.nochim,
                           file.path(ref_dir, "rdp_19_toSpecies_trainset.fa.gz"),
                           multithread = TRUE, tryRC = TRUE)

dir.create("results/taxa/rdp", showWarnings = FALSE, recursive = TRUE)
saveRDS(taxa.rdp, "results/taxa/rdp/taxa_species.rds")
head(taxa.rdp)
```

### Post-filter taxonomy (optional)

Matches `scripts/postfilter_taxa.R`: drop non-Bacteria/Archaea and
mitochondrial/chloroplast ASVs (Kingdom filter using SILVA taxonomy), then
drop low-abundance and low-prevalence ASVs. Produces `seqtab.clean`
without overwriting `seqtab.nochim` — swap it in downstream (abundance
table, tree, phyloseq) if you want the filtered set.

```{r}
is_bacteria <- taxa[, "Kingdom"] %in% c("Bacteria", "Archaea") &
    !(taxa[, "Family"] %in% c("Mitochondria")) &
    !(taxa[, "Order"] %in% c("Chloroplast"))
is_bacteria[is.na(is_bacteria)] <- FALSE

seqtab.kingdom <- seqtab.nochim[, is_bacteria, drop = FALSE]
cat("Kingdom/organelle filter: kept", ncol(seqtab.kingdom), "of", ncol(seqtab.nochim), "ASVs\n")

MIN_ABUNDANCE <- 10
seqtab.abund <- seqtab.kingdom[, colSums(seqtab.kingdom) >= MIN_ABUNDANCE, drop = FALSE]
cat("Abundance>=", MIN_ABUNDANCE, " filter: kept ", ncol(seqtab.abund), " of ",
    ncol(seqtab.kingdom), " ASVs\n", sep = "")

seqtab.clean <- seqtab.abund[, colSums(seqtab.abund > 0) >= 2, drop = FALSE]
cat("Prevalence>=2 filter: kept", ncol(seqtab.clean), "of", ncol(seqtab.abund), "ASVs\n")
```

## Abundance table

Per-ASV counts, relative abundance, and taxonomy in one table — matches
`scripts/make_abundance_table.R`.

```{r}
counts <- t(seqtab.nochim)
shared <- intersect(rownames(counts), rownames(taxa))
counts.shared <- counts[shared, , drop = FALSE]
taxa.shared   <- taxa[shared, , drop = FALSE]

relabund <- sweep(counts.shared, 2, colSums(counts.shared), "/") * 100
colnames(relabund) <- paste0(colnames(counts.shared), "_pct")

abundance <- cbind(
    asv_id   = paste0("ASV", seq_len(nrow(counts.shared))),
    sequence = shared,
    as.data.frame(taxa.shared),
    as.data.frame(counts.shared),
    as.data.frame(relabund)
)

write.csv(abundance, "results/abundance_table_silva.csv", row.names = FALSE)
head(abundance)
```

## Taxonomic composition plot

Matches `scripts/plot_composition.R`: aggregates genus-level relative
abundance by V-region and plots a stacked bar chart. Needs an SRA-style
metadata CSV with `Run` and `Sample Name` columns (the latter encoding the
V-region, e.g. `..._V3-V4`) — only meaningful when a study spans multiple
V-regions per sample set. Left `eval=FALSE`; flip to `eval=TRUE` once
`meta_csv` points at a real file.

```{r composition-plot, eval=FALSE}
top_n    <- 20
meta_csv <- "sra_metadata.csv"

meta    <- read.csv(meta_csv, check.names = FALSE, stringsAsFactors = FALSE)
srr_col  <- "Run"
name_col <- "Sample Name"

extract_vregion <- function(x) {
    m <- regmatches(x, regexpr("V\\d+(-V\\d+)*$", x))
    ifelse(length(m) == 0, x, m)
}
meta$vregion <- sapply(meta[[name_col]], extract_vregion)
srr_to_vregion <- setNames(meta$vregion, meta[[srr_col]])

common <- intersect(rownames(seqtab.nochim), names(srr_to_vregion))
seqtab.plot <- seqtab.nochim[common, , drop = FALSE]
group <- srr_to_vregion[common]

taxa_df <- as.data.frame(taxa)
genus <- ifelse(
    !is.na(taxa_df$Genus),    taxa_df$Genus,
    ifelse(!is.na(taxa_df$Family), paste0("unknown_", taxa_df$Family),
    ifelse(!is.na(taxa_df$Order),  paste0("unknown_", taxa_df$Order),
                                   "unassigned"))
)
names(genus) <- rownames(taxa_df)

counts <- t(seqtab.plot)
shared <- intersect(rownames(counts), names(genus))
counts <- counts[shared, , drop = FALSE]
genus  <- genus[shared]

vregions   <- sort(unique(group))
agg_counts <- sapply(vregions, function(vr) {
    cols <- names(group)[group == vr]
    cols <- intersect(cols, colnames(counts))
    if (length(cols) == 1) counts[, cols]
    else rowSums(counts[, cols, drop = FALSE])
})
colnames(agg_counts) <- vregions

genus_counts <- rowsum(agg_counts, group = genus)
relabund     <- sweep(genus_counts, 2, colSums(genus_counts), "/") * 100

sorted   <- sort(rowSums(relabund), decreasing = TRUE)
top_taxa <- names(sorted)[seq_len(min(top_n, length(sorted)))]
top_taxa <- top_taxa[top_taxa %in% rownames(relabund)]

rest  <- rownames(relabund)[!rownames(relabund) %in% top_taxa]
other <- if (length(rest) > 0) colSums(relabund[rest, , drop = FALSE]) else rep(0, ncol(relabund))

plot_mat <- rbind(relabund[top_taxa, , drop = FALSE], other = other)
df       <- as.data.frame(plot_mat)
df$taxon <- factor(rownames(df), levels = rev(c(top_taxa, "other")))
df_long  <- pivot_longer(df, cols = -taxon, names_to = "vregion", values_to = "pct")
df_long$vregion <- factor(df_long$vregion, levels = vregions)

comp_plot <- ggplot(df_long, aes(x = vregion, y = pct, fill = taxon)) +
    geom_bar(stat = "identity", width = 0.8) +
    scale_y_continuous(expand = c(0, 0), limits = c(0, 100)) +
    labs(x = NULL, y = "Relative abundance (%)", fill = NULL) +
    theme_bw(base_size = 11) +
    theme(
        axis.text.x        = element_text(angle = 45, hjust = 1),
        legend.position    = "right",
        panel.grid.major.x = element_blank()
    )

ggsave("results/plots/composition_silva.pdf", comp_plot, width = 14, height = 7)
comp_plot
```

## Phylogenetic tree (DECIPHER + ggtree)

Same approach as `scripts/build_tree.R` (DECIPHER `AlignSeqs` + `Treeline`,
NJ under a GTR model) so it only needs the packages already declared in
`envs/dada2.yaml` (no `phangorn` dependency).

```{r message=FALSE, warning=FALSE}
# Pull ASV sequences from the chimera-free table
seqs <- getSequences(seqtab.nochim)
names(seqs) <- seqs   # required downstream for tip labeling

# Multiple sequence alignment
alignment <- AlignSeqs(DNAStringSet(seqs),
                       anchor = NA,
                       processors = NULL,
                       verbose = FALSE)

# NJ tree under a GTR model
tree_dend <- Treeline(alignment,
                      method  = "NJ",
                      model   = "GTR",
                      verbose = FALSE)

asv.tree <- as.phylo(as.hclust(tree_dend))

# Save the tree for reuse
dir.create("results/tree", showWarnings = FALSE, recursive = TRUE)
saveRDS(asv.tree, "results/tree/tree.rds")
write.tree(asv.tree, "results/tree/tree.nwk")
```

Visualize the tree with taxonomy-labeled tips (Genus + Species where
available, colored by Phylum):

```{r fig.width=12, fig.height=14, message=FALSE, warning=FALSE}
tree_taxa <- as.data.frame(taxa)
tree_taxa <- tree_taxa[, !duplicated(colnames(tree_taxa)), drop = FALSE]
tree_taxa <- cbind(label = rownames(tree_taxa), tree_taxa)

tree_taxa$tip_label <- ifelse(
    !is.na(tree_taxa$Species), paste(tree_taxa$Genus, tree_taxa$Species),
    ifelse(!is.na(tree_taxa$Genus), tree_taxa$Genus,
    ifelse(!is.na(tree_taxa$Family), tree_taxa$Family, tree_taxa$Order)))

tree_plot <- ggtree(asv.tree) %<+% tree_taxa +
    geom_tiplab(aes(label = tip_label, color = Phylum), size = 2.5, align = TRUE) +
    theme_tree2() +
    theme(legend.position = "bottom")

ggsave("results/tree/tree.pdf", tree_plot, width = 12, height = 14)
tree_plot
```

## Build the phyloseq object (with tree)

Prepare metadata (QIIME2-style `sample-metadata.tsv` + `manifest.tsv`).

```{r}
metadata <- read.table(file = "sample-metadata.tsv",
                       comment.char = "", header = TRUE, sep = "\t")
manifest <- read.table(file = "manifest.tsv",
                       header = TRUE, sep = "\t")

manifest$absolute.filepath <- substring(manifest$absolute.filepath, 6)

merged_table   <- merge(manifest, metadata,
                        by.x = "sample.id", by.y = "X.SampleID")
resulting_table <- merged_table[, c(2, 1, 3:7)]
rownames(resulting_table) <- merged_table$absolute.filepath
resulting_table <- resulting_table[, -1]
```

Assemble the phyloseq object, including the tree.

```{r}
ps <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows = FALSE),
               sample_data(resulting_table),
               tax_table(taxa),
               phy_tree(asv.tree))
```

Replace long DNA strings with short ASV IDs (DNA preserved in `refseq`).

```{r}
dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps <- merge_phyloseq(ps, dna)
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))
ps
```

## Phylogenetic diversity (optional)

Now phylogenetic metrics are available, e.g.:

```{r eval=FALSE}
# Faith's PD via picante (if installed)
# library(picante)
# pd <- picante::pd(as(otu_table(ps), "matrix"), phy_tree(ps))

# UniFrac distances
unifrac.uw <- UniFrac(ps, weighted = FALSE)
unifrac.w  <- UniFrac(ps, weighted = TRUE)
```
